---
title: "Available Color Palettes"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| echo: false
#| include: false
library(scales)

# Built the total set of palettes 
palettes_hcl <- rbind(
    data.frame(
      type = "qual", 
      family = "hcl", 
      palette = grDevices::hcl.pals("qualitative")
    ),
    data.frame(
      type = "seq",
      family = "hcl",
      palette = grDevices::hcl.pals(type = "sequential")
    ),
    data.frame(
      type = "div",
      family = "hcl",
      palette = grDevices::hcl.pals(type = "diverging")
    )
  )

palettes_base <- utils::getFromNamespace("palette.pals", "grDevices")() |> 
  data.frame(type = "qual", family = "base", palette = _)

palettes_viridis <- c(
    "magma",
    "inferno",
    "plasma",
    "viridis",
    "cividis",
    "rocket",
    "mako",
    "turbo"
  ) |> 
    data.frame(type = "seq", palette = _, family = "viridis")

palettes_brewer <- rownames(RColorBrewer::brewer.pal.info) |> 
  data.frame(type = RColorBrewer::brewer.pal.info$category, family = "brewer", palette = _)

palettes_dichromat <- names(dichromat::colorschemes) |> 
  data.frame(family = "dichromat", palette = _)
  
palettes_dichromat$type <- "div"
palettes_dichromat$type[palettes_dichromat$palette == "Categorical.12"] <- "qual"
palettes_dichromat$type[
  palettes_dichromat$palette %in% c(
    "SteppedSequential.5", "LightBluetoDarkBlue.10", "LightBluetoDarkBlue.7")] <- "seq"

palettes_dichromat <- palettes_dichromat[c("type", "family", "palette")]

palettes_ggplot2 <- data.frame(
  family = "ggplot2",
  type = c("seq", "qual"),
  palette = c("grey", "hue")
)

# combine all the palettes together and overwrite any that are later repeated
palettes_all <- rbind(
  palettes_hcl,
  palettes_base,
  palettes_viridis,
  palettes_brewer,
  palettes_dichromat,
  palettes_ggplot2)
 
palettes_all$palette_lower <- tolower(palettes_all$palette)

palettes_all <- palettes_all[!duplicated(palettes_all$palette_lower, fromLast = TRUE),]
palettes_all <- palettes_all[palettes_all$palette_lower %in% scales:::palette_names(),]

palettes_all$type <- factor(palettes_all$type, levels = c("qual", "seq", "div"), ordered = TRUE)

palettes_all <- palettes_all[order(palettes_all$type, palettes_all$family, palettes_all$palette),]

```

{scales} makes `{r} length(scales:::palette_names())` color palettes available to packages like {ggplot2}, which references them using the `palette` argument of `scale_*_discrete()`, `scale_*_continuous()`, and `scale_*_binned()` functions.

This article groups every built-in name by the use it's best suited for: as a **discrete** palette for unordered categories or as a **continuous** palette for magnitude data (sequential or diverging).

```{r}
#| echo: false
# helper functions to display color bars
display_brewer <- function(palette_df) {
  brewer_info <- RColorBrewer::brewer.pal.info
  brewer_info$palette <- rownames(brewer_info)

  palette_df <- merge(
    brewer_info,
    palette_df[c("palette", "type")],
    by.x = c("palette", "category"),
    by.y = c("palette", "type")
  )

  old_par <- par(mfrow = c(nrow(palette_df), 1), mar = c(0.1, 5.1, 0.1, 1.1))
  for (i in 1:nrow(palette_df)) {
    cols <- pal_brewer(type = palette_df$category[i], palette = palette_df$palette[i])(palette_df$maxcolors[i])
    image(matrix(seq_along(cols)), col = cols, axes = FALSE)
    mtext(palette_df$palette[i], side = 2, las = 1, line = 1, cex = 0.8)
  }
  par(old_par)
}

display_hcl <- function(names, n = 100, discrete = FALSE) {
  if (discrete) n <- 8
  old_par <- par(mfrow = c(length(names), 1), mar = c(0.1, 5.6, 0.1, 1.1))
  for (nm in names) {
    cols <- grDevices::hcl.colors(n, palette = nm)
    image(matrix(seq_len(n)), col = cols, axes = FALSE)
    mtext(nm, side = 2, las = 1, line = 1, cex = 0.6)
  }
  par(old_par)
}

display_fixed <- function(cols_list, mar_left = 6.5, cex = 0.6) {
  old_par <- par(mfrow = c(length(cols_list), 1), mar = c(0.1, mar_left, 0.1, 1.1))
  for (nm in names(cols_list)) {
    cols <- cols_list[[nm]]
    image(matrix(seq_along(cols)), col = cols, axes = FALSE)
    mtext(nm, side = 2, las = 1, line = 1, cex = cex)
  }
  par(old_par)
}

# Make palette names easier for users to get
palette_names <- function(type) {
  x <- palettes_all$palette[palettes_all$type == type]
  paste(sprintf('`"%s"`', x), collapse = ", ")  
}
```

<details>
<summary>Palette names</summary>

```{r}
#| echo: false
#| results: asis

cat("**Discrete:** ", paste(sprintf('`"%s"`', palettes_all$palette[palettes_all$type == "qual"]), collapse = ", "), "\n")
```

```{r}
#| echo: false
#| results: asis
cat("**Sequential:** ", paste(sprintf('`"%s"`', palettes_all$palette[palettes_all$type == "seq"]), collapse = ", "), "\n")
```

```{r}
#| echo: false
#| results: asis
cat("**Diverging:** ", paste(sprintf('`"%s"`', palettes_all$palette[palettes_all$type == "div"]), collapse = ", "), "\n")
```

</details>

## Discrete palettes

Discrete palettes assign a distinct color to each category, e.g. `scale_color_discrete(palette = "Set2")`. These come from four sources: HCL's qualitative palettes, base R's fixed categorical sets, Brewer's qualitative palettes, and one dichromat palette built specifically for categories (plus ggplot2's own default, `"hue"`). Where a name is shared across sources, the last-registered palette name takes precedence; for instance, `"Blues"` exists in both the HCL and Brewer families, but {scales} only returns the Brewer version of `"Blues"`.

Brewer's qualitative palettes:

```{r}
#| echo: false
#| fig-height: 2.67
palettes_all[palettes_all$family == "brewer" & palettes_all$type == "qual", ] |> 
  display_brewer()
```

Base R's fixed categorical sets (`grDevices::palette.pals()`, R >= 4.0.0 only), including familiar defaults like `"R3"` and `"R4"`, alongside color-vision-deficiency-aware sets such as `"Okabe-Ito"`:

```{r}
#| echo: false
#| fig-height: 4.33
base_names <- utils::getFromNamespace("palette.pals", "grDevices")() |> 
  intersect(palettes_all[palettes_all$family == "base",]$palette) 

base_names <- base_names[!grepl("ggplot2", base_names)]

palette_fn <- utils::getFromNamespace("palette.colors", "grDevices")
base_cols <- setNames(lapply(base_names, function(nm) palette_fn(palette = nm)), base_names)

display_fixed(base_cols)
```

HCL's qualitative palettes, generated from a continuous hue rotation rather than a fixed set of hand-picked colors (shown here at `n = 8`; requesting many more levels than that can produce hues that are difficult to distinguish):

```{r}
#| echo: false
#| fig-height: 1.67
palettes_all[palettes_all$family == "hcl" & palettes_all$type == "qual", ]$palette |> 
  display_hcl(discrete = TRUE)
```

One more standalone discrete palette: ggplot2's default `"hue"` palette:

```{r}
#| echo: false
#| fig-height: 0.33
display_fixed(
  list(
    hue = pal_hue()(8)
  ),
  mar_left = 2,
  cex = 0.65
)
```

Dichromat's single qualitative palette:

```{r}
#| echo: false
#| fig-height: 0.33
palettes_all[palettes_all$type == "qual" & palettes_all$family == "dichromat", ]$palette |> 
  {\(x) dichromat::colorschemes[x]}() |> 
  display_fixed(mar_left = 4.5, cex = 0.55)
```

Many of these palettes have a fixed, limited number of colors. Requesting more colors than a palette contains will recycle values (for the fixed sets) or produce increasingly similar hues (for HCL's qualitative palettes and for ggplot2's `"hue"` palette), so these are best suited to categorical data with a known, modest number of levels.

## Continuous palettes

Continuous palettes map a numeric range onto a smooth color gradient, e.g. `scale_color_continuous(palette = "viridis")`. These split naturally into two subtypes: **sequential** palettes for data that increases in magnitude from a natural minimum, and **diverging** palettes for data that departs from a meaningful midpoint (often zero). Sequential and diverging palettes can also be used with binned scales, which quantize the continuous palette into discrete steps.

### Sequential

The viridis family: perceptually uniform, color-blind friendly gradients, and a strong general-purpose default:

```{r}
#| echo: false
#| fig-height: 2.67
opts <- c("magma", "inferno", "plasma", "viridis", "cividis", "rocket", "mako", "turbo")
old_par <- par(mfrow = c(length(opts), 1), mar = c(0.1, 4.6, 0.1, 1.1))
for (opt in opts) {
  cols <- pal_viridis(option = opt)(256)
  image(matrix(seq_along(cols)), col = cols, axes = FALSE)
  mtext(opt, side = 2, las = 1, line = 1, cex = 0.8)
}
par(old_par)
```

Brewer's sequential palettes:

```{r}
#| echo: false
#| fig-height: 6
palettes_all[palettes_all$type == "seq" & palettes_all$family == "brewer",] |> 
  display_brewer()
```

HCL's sequential palettes — by far the largest family, including several scientific color maps (`"Batlow"`, `"Hawaii"`, `"Turku"`, `"Lajolla"`) alongside classics like `"Heat"` and `"Terrain"`:

```{r}
#| echo: false
#| fig-height: 16
palettes_all[palettes_all$type == "seq" & palettes_all$family == "hcl",]$palette |> 
  display_hcl()
```

Dichromat's two color-safe ramps and its unique stepped-sequential ramp:

```{r}
#| echo: false
#| fig-height: 1
palettes_all[palettes_all$type == "seq" & palettes_all$family == "dichromat",]$palette |> 
  {\(x) dichromat::colorschemes[x]}() |> 
  display_fixed(mar_left = 9, cex = 0.55)
```

And a plain greyscale ramp, useful whenever color needs to be avoided or reserved for another variable:

```{r}
#| echo: false
#| fig-height: .33
display_fixed(list(grey = grDevices::grey.colors(100)), mar_left = 2.75, cex = 0.8)
```

### Diverging

Brewer's diverging palettes:

```{r}
#| echo: false
#| fig-height: 3
palettes_all[palettes_all$type == "div" & palettes_all$family == "brewer",] |> 
  display_brewer()
```

HCL's diverging palettes:

```{r}
#| echo: false
#| fig-height: 6
palettes_all[palettes_all$type == "div" & palettes_all$family == "hcl",]$palette |> 
  display_hcl()
```

Dichromat's color-blind-safe diverging palettes:

```{r}
#| echo: false
#| fig-height: 4.33
palettes_all[palettes_all$type == "div" & palettes_all$family == "dichromat",]$palette |> 
  {\(x) dichromat::colorschemes[x]}() |> 
  display_fixed(mar_left = 8, cex = 0.55)
```
